---
title: "Enrichment analysis (EA)"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
    number_sections: true
    df_print: paged
params:
  data: NA
  deaanalysistype: NA
  ea_plottype: NA
  gsea_gene_sets: NA
  enrichmentfdr: NA
  msigdbtype: NA
  gotype: NA
  deaanalysisselect: NA
  ea_nb_categories: NA
  earankingmethod: NA
  ea_subsetfdr: NA
  ea_subsetlc: NA
  ea_subsettype: NA
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,fig.width = 10,fig.height = 10)
```


# Load required libraries
```{r libraries, message = FALSE}
library(plotly)
library(plyr)
library(dplyr)
```

# Input DEA file
```{R}
dea_result <- params$data$dea.results
dea_result %>% head
```


## Add gene information

### Retrieve gene information

```{R biomart, eval = FALSE}
library(biomaRt)
attributes <-  c(
  "chromosome_name",
  "start_position",
  "end_position",
  "strand",
  "ensembl_gene_id",
  "entrezgene_id",
  "external_gene_name"
)

# Get last version of human from EMSEMBL using  biomaRt
human <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
hg38 <- getBM(attributes =  attributes, mart = human)
```

```{R biomartlocal,include = FALSE, eval = TRUE}
hg38 <- params$data$hg38
hg38$entrezgene_id <- hg38$entrezgene
```

```{R mapgenes}
dea_result$entrezgene_id <- hg38$entrezgene_id[match(dea_result$Symbol,hg38$external_gene_name)]
dea_result$ensembl_gene_id <- hg38$ensembl_gene_id[match(dea_result$Symbol,hg38$external_gene_name)]
head(dea_result)
```

# Selecting genes from DEA file

## GSEA analysis - Ranking method

Observation: if method is `ORA`, the ranking will not be used, only the gene names will be.

```{r gsearankmethod, eval=TRUE}
# Calculate method score for each gene using the DEA file
params$earankingmethod
if(params$earankingmethod == "log Fold Change") {
  geneList.metric <- dea_result$log2FoldChange
} else if(params$earankingmethod ==  "-log10(P-value) * sig(log2FC)") {
  geneList.metric <- -log10(dea_result$pvalue) * sign(dea_result$log2FoldChange)
} else {
  geneList.metric <- -log10(dea_result$pvalue) * dea_result$log2FoldChange
}
```

```{r gsearankingensembl, eval = (params$deaanalysisselect == "Gene Ontology Analysis"), include =   (params$deaanalysisselect == "Gene Ontology Analysis" )}
# Gene entrez ENSEMBL version (used in Gene Ontology Analysis")
geneList <- geneList.metric
names(geneList) <- dea_result$ensembl_gene_id
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]
```

```{r gsearankingentrez, eval=(params$deaanalysisselect != "Gene Ontology Analysis"), include = (params$deaanalysisselect != "Gene Ontology Analysis" )}
# Gene entrez ID version   
geneList <- geneList.metric 
names(geneList) <- dea_result$entrezgene_id
geneList <- sort(geneList, decreasing = TRUE)
geneList <- geneList[!is.na(names(geneList))]
```

`r if (params$deaanalysistype != "ORA")'## ORA analysis - Selecting significant genes'`

```{r oraparams, eval= (params$deaanalysisselect != "Gene Ontology Analysis"  & params$deaanalysistype == "ORA"), include = (params$deaanalysisselect != "Gene Ontology Analysis"  & params$deaanalysistype == "ORA")}
params$ea_subsettype # Direction (upregulated or downregulated genes)
params$ea_subsetlc # log FC cut-off
params$ea_subsetfdr # FDR cut-off
```

```{r oraentrez, eval= (params$deaanalysisselect != "Gene Ontology Analysis"  & params$deaanalysistype == "ORA"), include = (params$deaanalysisselect != "Gene Ontology Analysis"  & params$deaanalysistype == "ORA")}
ora <- dea_result[abs(dea_result$log2FoldChange) > params$ea_subsetlc & dea_result$pvalue < params$ea_subsetfdr,]
if(params$ea_subsettype == "Upregulated"){
  ora <- ora[ora$log2FoldChange > 0,]
} else {
  ora <- ora[ora$log2FoldChange < 0,]
}
dea.genes <- na.omit(ora$entrezgene_id)
```

```{r oraensembl, eval=(params$deaanalysisselect == "Gene Ontology Analysis"  & params$deaanalysistype == "ORA"), include = (params$deaanalysisselect == "Gene Ontology Analysis"  & params$deaanalysistype == "ORA")}
ora <- dea_result[abs(dea_result$log2FoldChange) > params$ea_subsetlc & dea_result$pvalue < params$ea_subsetfdr,]
if(params$ea_subsettype == "Upregulated"){
  ora <- ora[ora$log2FoldChange > 0,]
} else {
  ora <- ora[ora$log2FoldChange < 0,]
}
dea.genes <- na.omit(ora$ensembl_gene_id)
```

### Evaluated genes
```{R genes, eval=( params$deaanalysistype == "ORA"), include = (params$deaanalysistype == "ORA")} 
head(dea.genes)
length(dea.genes)
```

### Background genes
```{R background_genes}
head(geneList)
length(geneList)
```    

## Enrichment analysis (EA) parameters
```{r eaparams}
params$deaanalysistype
params$deaanalysisselect
params$ea_plottype
params$enrichmentfdr
```

# Enrichment analysis (EA)

`r if (params$deaanalysisselect  ==  "WikiPathways analysis")'## WikiPathways analysis'`

```{R WikiPathways, eval = (params$deaanalysisselect ==  "WikiPathways analysis"), include =   (params$deaanalysisselect ==  "WikiPathways analysis")}

wpgmtfile <- system.file("extdata/wikipathways-20180810-gmt-Homo_sapiens.gmt", 
                         package = "clusterProfiler")
wp2gene <- read.gmt(wpgmtfile)
wp2gene <- wp2gene %>% tidyr::separate(ont, c("name","version","wpid","org"), "%")
wpid2gene <- wp2gene %>% dplyr::select(wpid, gene) #TERM2GENE
wpid2name <- wp2gene %>% dplyr::select(wpid, name) #TERM2NAME

if(params$deaanalysistype == "ORA"){
  results <- enricher(dea.genes, 
                      TERM2GENE = wpid2gene, 
                      universe  = names(geneList),
                      TERM2NAME = wpid2name,
                      pvalueCutoff = params$enrichmentfdr
  )
} else {
  results <- GSEA(geneList, 
                  TERM2GENE = wpid2gene, 
                  TERM2NAME = wpid2name, 
                  verbose = FALSE,
                  pvalueCutoff = params$enrichmentfdr
  )
}
```


`r if (params$deaanalysisselect == "MSigDb analysis")'##  MSigDb analysis'`

```{R MSigDb, eval = (params$deaanalysisselect == "MSigDb analysis"), include = (params$deaanalysisselect == "MSigDb analysis")}
params$msigdbtype
if(params$msigdbtype != "All"){
  m_t2g <- msigdbr(species = "Homo sapiens", category = params$msigdbtype) %>% 
    dplyr::select(gs_name, entrez_gene)
} else {
  m_t2g <- msigdbr(species = "Homo sapiens") %>% 
    dplyr::select(gs_name, entrez_gene)
}
```


```{R MSigDbORA, eval = (params$deaanalysisselect == "MSigDb analysis"  & params$deaanalysistype == "ORA"), include = (params$deaanalysisselect == "MSigDb analysis"  & params$deaanalysistype == "ORA")}
results <- enricher(dea.genes, 
                    TERM2GENE = m_t2g,
                    universe = names(geneList),
                    pvalueCutoff = params$enrichmentfdr
)
```

```{R MSigDbGSEA, eval = (params$deaanalysistype != "ORA" & params$deaanalysisselect == "MSigDb analysis" ), include = (params$deaanalysistype != "ORA" & params$deaanalysisselect == "MSigDb analysis" )}
params$enrichmentfdr
results <- GSEA(geneList, 
                TERM2GENE = m_t2g,
                pvalueCutoff =  params$enrichmentfdr
)
```

`r if (params$deaanalysisselect == "Gene Ontology Analysis")'##  Gene Ontology Analysis'`

```{R GOORA, eval = (params$deaanalysisselect == "Gene Ontology Analysis" & params$deaanalysistype == "ORA"), include = (params$deaanalysistype == "ORA" & params$deaanalysisselect == "Gene Ontology Analysis")}
params$gotype
results <- enrichGO(gene          = dea.genes,
                    universe      = names(geneList),
                    OrgDb         = org.Hs.eg.db,
                    keyType       = ifelse(all(grepl("ENSG",names(geneList))), "ENSEMBL","ENTREZID"),
                    ont           = params$gotype,
                    pAdjustMethod = "BH",
                    pvalueCutoff  = params$enrichmentfdr,
                    readable      = TRUE)
```

```{R GOGSEA, eval = (params$deaanalysisselect == "Gene Ontology Analysis" & params$deaanalysistype != "ORA"), include = (params$deaanalysistype != "ORA" & params$deaanalysisselect == "Gene Ontology Analysis")}
params$gotype
results <- gseGO(geneList     = geneList,
                 OrgDb        = org.Hs.eg.db,
                 ont          = params$gotype,
                 keyType      = ifelse(all(grepl("ENSG",names(geneList))), "ENSEMBL","ENTREZID"),
                 nPerm        = 1000,
                 minGSSize    = 100,
                 maxGSSize    = 500,
                 pvalueCutoff = params$enrichmentfdr,
                 verbose      = FALSE)
```


`r if (params$deaanalysisselect == "KEGG Analysis")'## KEGG Analysis'`

```{R KEGG, eval = (params$deaanalysisselect == "KEGG Analysis"), include = (params$deaanalysisselect == "KEGG Analysis")}
if(params$deaanalysistype == "ORA"){
  results <- enrichKEGG(dea.genes, 
                        pvalueCutoff = params$enrichmentfdr,
                        organism = "hsa")
} else {
  results <- gseKEGG(geneList, 
                     pvalueCutoff = params$enrichmentfdr,
                     organism = "hsa", 
                     nPerm = 10000)
}
```

`r if (params$deaanalysisselect == "Disease Ontology Analysis")'## Disease Ontology Analysis'`

```{R DOA, eval = (params$deaanalysisselect == "Disease Ontology Analysis"), include = (params$deaanalysisselect == "Disease Ontology Analysis")}
if(params$deaanalysistype == "ORA"){
  # Gene Ontology Analysis
  results <- enrichDO(gene          = dea.genes,
                      universe      = names(geneList),
                      pAdjustMethod = "BH",
                      ont           = "DO",
                      pvalueCutoff  = params$enrichmentfdr,
                      readable      = TRUE)
  
} else {
  results <- gseDO(geneList     = geneList,
                   nPerm        = 1000,
                   minGSSize    = 10,
                   maxGSSize    = 500,
                   pvalueCutoff = params$enrichmentfdr,
                   verbose      = FALSE)
}
```


`r if (nrow(summary(results)) > 0)'## Results table'`

```{R table, eval = (nrow(summary(results)) > 0), include = (nrow(summary(results)) > 0)}
results %>% summary %>% DT::datatable(options = list(scrollX = TRUE,  class = 'cell-border stripe'))
```


`r if (nrow(summary(results)) > 0)'## Plots'`

`r if (params$deaanalysistype == "ORA" & nrow(summary(results)) > 0)'### Dot plot'`

```{R ORA_dot, eval = (params$deaanalysistype == "ORA"  & nrow(summary(results)) > 0), include = (params$deaanalysistype == "ORA" & nrow(summary(results)) > 0)}
params$ea_nb_categories
dotplot(results, showCategory = params$ea_nb_categories)
```


`r if (params$deaanalysistype == "ORA" & nrow(summary(results)) > 0)'### Enrichment map (network)'`

```{R ORA_map, eval = (params$deaanalysistype == "ORA" & nrow(summary(results)) > 0), include = (params$deaanalysistype == "ORA" & nrow(summary(results)) > 0)}
emapplot(results)
```

`r if (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)'### Dot plot'`

```{R, eval = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0), include = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)}
params$ea_nb_categories
dotplot(results, showCategory = params$ea_nb_categories) + 
  facet_grid(.~ifelse(NES < 0, 'suppressed', 'activated'))
```

`r if (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)'### Ridgeline'`

```{R, eval = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0), include = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)}
params$ea_nb_categories
ridgeplot(results,showCategory = params$ea_nb_categories)
```

`r if (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)'### Running score and preranked list'`

```{R, eval = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0), include = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)}
gseaplot2(results, geneSetID = match(params$gsea_gene_sets, results$Description))
```

`r if (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)'### Ranked list of genes'`

```{R, eval = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0), include = (params$deaanalysistype != "ORA" & nrow(summary(results)) > 0)}
p <- gsearank(results,  which(params$gsea_gene_sets ==  results$Description), 
              title = results[ match(params$gsea_gene_sets, results$Description), "Description"])
params$gsea_gene_sets
pp <- lapply( match(params$gsea_gene_sets, results$Description), function(i) {
  anno <- results[i, c("NES",  "p.adjust")]
  lab <- paste0(names(anno), "=",  round(anno, 3), collapse = "\n")
  
  es <- results[i, "enrichmentScore"]
  x <- ifelse(es < 0, 0, length(geneList) * .8)
  gsearank(results, i, results[i, 2]) + 
    xlab(NULL) +
    ylab(NULL) +
    annotate("text", 
             x,  
             es * .5, 
             label = lab, 
             hjust = 0, 
             vjust = 0, 
             size = 4) + xlim(0, 12500)
  
})
p <- plot_grid(plotlist = pp, ncol = 1)

p
```

# Session Information
```{r sessionInfo}
sessionInfo()
```
